Discrete Ordinates

### 2017-06-01 docprep/chapmeth.tex
Reworded sections of the methods and fixed typos in the intro

This chapter summarizes the solution methodology employed by DOCTORS. The techniques used to compute the collided and uncollided fluxes are described in seperate sections. After the flux solution methodology is explained, the methods used for source generation and flux-to-dose conversion are covered.

@@

The linear Boltzmann equation (LBE) is generally true of any particle given that outside forces (electromagnetic or graitational) are either not present or negligible and particle-particle collisions are insignificant. Therefore, the LBE is generally not valid for charged parpticles or systems whose particle density is on the order of the medium they are traveling trhrough. In such systems, particle-particle ineractions may not be negligible.

@@

\begin{equation} \label{eq:boltz_t_dep}
\begin{split}
	&\left[ \frac{1}{v(\boldsymbol{r},t)} \frac{\partial}{\partial t} + \hat{\Omega} \cdot \nabla + \Sigma_t(\boldsymbol{r}, E, t) \right]
	\psi(\boldsymbol{r}, E, \hat{\Omega}, t) = \\
	&\int_{4 \pi} \int_0^\infty \Sigma_s(\boldsymbol{r}, E' \rightarrow E, \hat{\Omega}' \rightarrow \hat{\Omega}, t) \psi(\boldsymbol{r}, E', \hat{\Omega}', t) dE' d\hat{\Omega}' + S(\boldsymbol{r}, E, \hat{\Omega}, t)
\end{split}
\end{equation}

\noindent
where $v$ is the velocity of the photon in the medium at $\boldsymbol{r}$, $\Sigma_t$ is the total macroscopic cross section, $\Sigma_s$ is the macroscopic scattering cross section, $S$ is an external photon source, and $\psi$ is the total photon angular flux.

The transient case is not typically of interest in medical diagnostic imaging, rather the steady state is more important. The time dependence in Eq. \ref{eq:boltz_t_dep} is removed to yield

\begin{equation} \label{eq:boltz}
\begin{split}
	&\left[ \hat{\Omega} \cdot \nabla + \Sigma_t(\boldsymbol{r}, E) \right]
	\psi(\boldsymbol{r}, E, \hat{\Omega}) = \\
	&\int_{4 \pi} \int_0^\infty \Sigma_s(\boldsymbol{r}, E' \rightarrow E, \hat{\Omega}' \rightarrow \hat{\Omega}) \psi(\boldsymbol{r}, E', \hat{\Omega}') dE' d\hat{\Omega}' + S(\boldsymbol{r}, E, \hat{\Omega})
\end{split}
\end{equation}

The steady state form of the LBE can be readily derived from simple intuition. In the steady state, particle production and removal must be equal. In any volume, three production mechanisms are present. Particles can stream into the volume across a bounding surface, they can inscatter from another energy-direction component of phase space, or they can be produced directly by the external source. Three removal terms are also present, particles can stream out of the volume of interest by crossing a bounding surface, be absorbed and vanish entirely, or outscatter to a energy-direction componennt of phase space other than the one of interest.

The downscatter and absorbtiption removal terms are typically grouped together as

\begin{equation}
\Sigma_t(\boldsymbol{r}, E) \psi(\boldsymbol{r}, E, \hat{\Omega})
\end{equation}

\noindent
and both production and removal streaming operators are combined as well as

\begin{equation}
\hat{\Omega} \cdot \nabla \psi(\boldsymbol{r}, E, \hat{\Omega})
\end{equation}

\noindent
and included as a removal term since they normal is defined pointing outward from the volume. The normal sign convention results in particles streaming out being positive and those streaming in beocming negative. The inscatter to a particular energy and direction is computed by integrating over all other energies and diretions

\begin{equation}
\int_{4\pi}^{} \int_{0}^{\infty} \Sigma_s(\boldsymbol{r}, E' \rightarrow E, \hat{\Omega}' \rightarrow \hat{\Omega}) \psi(\boldsymbol{r}, E', \hat{\Omega}') dE' d\hat{\Omega}'
\end{equation}

The following sections show how Eq.~\ref{eq:boltz} is discretized and solved computationally.

\subsection{Angular Discretization}
The coordinate system used in DOCTORS is shown in Fig.~\ref{fig:coord_sys}. A discrete set of $N_a$ angles ($\Omega_{a}, \quad a = 0 \ldots N_a-1$) is selected to represent continuous directional space. Particles are transported only along these discrete directions from a voxel to adjacent voxels.

@@

THEORY

To inlcude a user defined quadrature, the plain text file should be formatted such that it contains four columns similar to Table~\ref{tab:quad_format} (without any headings). No checks are performed to ensure that octants are covered, this allows simple testing quadratures that utilize only a single direction to be used for testing purposes. Values for $\mu$, $\eta$, and $\xi$ do not necessarily need to add to unity, they will be automatically normalized by DOCTORS. Likewise, the weights will be normalized such that their summation is unity. However, user defined directions \textit{should not} be parallel to any major axis. Doing so may result in undefined behavior. Also, after normalizaiton, no two rows should have identical values for $\mu$, $\eta$, and $\xi$ such as the last two rows of Table~\ref{tab:quad_interp}. Additionally, no weights may be negative.

@@

The data in Table~\ref{tab:quad_format} would be read in and renormalized by DOCTORS. The normalized data would be identical to the data in Table~\ref{tab:quad_interp}.

@@

A given direction, $\hat{\Omega}$, is determined by its three cosine components. This relation is given in Eq.~\ref{eq:omega_cos} where $\hat{i}$, $\hat{j}$, and $\hat{k}$ are the unit directions in the $x$, $y$, and $z$ directions respectively.

\begin{equation} \label{eq:omega_cos}
\hat{\Omega} = \mu \hat{i} + \eta \hat{j} + \xi \hat{k}
\end{equation}

Each discrete angle is associated a weight. The weights are designed with the angles such that once a discrete quadrature is selected, integration over continuous space becomes integration in discrete space approximated by

\begin{equation} \label{eq:disc_int}
\int_{4 \pi} f(\hat{\Omega}) d\hat{\Omega} \approx \sum_{a=0}^{N_a-1} f(\hat{\Omega}_a) \omega_a
\end{equation}

Discretizing Eq.~\ref{eq:boltz} in angular space gives

\begin{equation} \label{eq:boltz_a}
\begin{split}
&\left[ \hat{\Omega}_a \cdot \nabla + \Sigma_t(\boldsymbol{r}, E) \right]
\psi_{a}(\boldsymbol{r}, E) = \\
&\sum_{a=0}^{N_a-1} \int_0^\infty \Sigma_{s, a, a'}(\boldsymbol{r}, E' \rightarrow E) \psi_{a'}(\boldsymbol{r}, E') dE' \omega_a + S_a(\boldsymbol{r}, E)
\end{split}
\end{equation}

\noindent
where the subscript $a$ denotes the direction.

@@

Continuous energy is discretized into $G$ groups indexed from 0 to $G-1$. Upscatter is assumed to be negligible. Therefore, Eq.~\ref{eq:boltz_a} becomes the energy discretized Eq.~\ref{eq:boltz_e} where the group number is indexed by superscript $g$.

\begin{equation} \label{eq:boltz_e}
\left[ \hat{\Omega}_a \cdot \nabla + \Sigma_t^g(\boldsymbol{r}) \right]
\psi_{a}^{g}(\boldsymbol{r}) = 
\sum_{a=0}^{N_a-1} \sum_{g'=0}^{G-1} \Sigma_{s, a, a'}^{g, g'}(\boldsymbol{r}) \psi_{a'}^{g'}(\boldsymbol{r}) \omega_a + S_a^g(\boldsymbol{r})
\end{equation}

In photon based problems, upscatter is typicall negligible [CITE]. By the time photons become low enough in energy to upscatter significantly, they are no longer of dosimetric concern. Removing upscatter reduces Eq.~\ref{eq:boltz_e} to

\begin{equation} \label{eq:boltz_e2}
\left[ \hat{\Omega}_a \cdot \nabla + \Sigma_t^g(\boldsymbol{r}) \right]
\psi_{a}^{g}(\boldsymbol{r}) = 
\sum_{a=0}^{N_a-1} \sum_{g'=g}^{G-1} \Sigma_{s, a, a'}^{g, g'}(\boldsymbol{r}) \psi_{a'}^{g'}(\boldsymbol{r}) \omega_a + S_a^g(\boldsymbol{r})
\end{equation}

\noindent
which is very similar; only the starting point for the scatter summation changes. Computationall, though, this makes a significant different. This means that the highest energy group can be solved independently of all those below it. Once that group is solved, the next group can also be solved and so forth.

@@

The group structure cannot be arbitrarily selected by DOCOTRS, instead it is determined by the external cross section data file. The energy group structure should be selected appropriately for the problem. The cross sections currently available are those available from the SCALE 6.2 distribution. Those cross sections are designed for light water reactor analysis and are not well suited to medical physics applications. This is currently one of the greatest limitaitons of DOCOTRS. The SCALE cross section data also contains neutron data which is useless for the CT dosimetry DOCTORS performs. More details of the cross section parsing and generation are included in Section~\ref{sec:xsparse} and~\ref{sec:xsgen}.

Once the energy group structure is selected, cross section data must be available. Group averaged cross section values from energy $E_1$ to $E_2$ are computed using Eq.~\ref{eq:groupxs} where $\sigma(E)$ is the continuous microscopic cross section. A weighting function, $f$, is used to weight some energies. The weighting function can be any of a number commonly used functions.

\begin{equation}\label{eq:groupxs}
\sigma_G = \frac{\int_{E_1}^{E_2}f(E')\sigma(E') dE'}{\int_{E_1}^{E_2} dE'}
\end{equation}

@@

The problem domain is split into an evenly spaced Cartesian grid. The grid spacing between each of the spatial dimensions does not necessarily need to be uniform, but it often is in the case of CT voxel phantoms.

The problem domain must be a regular, rectangular parallelpiped of dimension $D_x \times D_y \times D_z$. The mesh is partitioned into $N_x$, $N_y$, and $N_z$ evenly spaced bins along each major direction. The total number of voxels, $N_V$ is easily computed as the multiple of each of the component directions given in Eq.~\ref{eq:n_v}.

\begin{equation} \label{eq:n_v}
	N_V = N_x N_y N_z
\end{equation}

Voxels can be indexed in one of two ways, eitehr be their component indices or its global index. Each voxel in the mesh has three components indices, one in each major direction, $ix$, $i_y$, and $i_z$ that rante from 0 to $N_x - 1$, $N_y - 1$, or $N_z-1$. However, this makes indexing cumbersome, so they are combined into a unique glorbal index, $i$ using Eq.~\ref{eq:n_v}. In general, this is known as "flattening" a matrix into a one-dimensional vector.

@@

The spatial mesh is shown in Fig.~\ref{fig:spatial_disc}. The length of a voxel along a major direction is computed by dividing the size of the problem domain along that dimension by the number of mesh elements. Eq.~\ref{eq:mesh_x} shows this computation for the $x$ direction. Analagous equations apply in the $y$ and $z$ directions as well.

\begin{equation} \label{eq:mesh_x}
\Delta x = \frac{D_x}{N_x}
\end{equation}

The fully discretized form of the steady state LBE given in Eq.~\ref{eq:boltz} is

\begin{equation} \label{eq:boltz_i}
\left[ \hat{\Omega}_a \cdot \nabla + \Sigma_{t,i}^g \right]
\psi_{i,a}^{g} = 
\sum_{a=0}^{N_a-1} \sum_{g'=g}^{G-1} \Sigma_{s, i, a, a'}^{g, g'} \psi_{i, a'}^{g'} \omega_a + S_{i,a}^g
\end{equation}

@@

To solve the fully discretized form of the Linear Boltzmann Equation given in Eq.~\ref{eq:boltz_i}, the gradient operator must first be computed numerically. Recall that $\Omega_a$ can be written in vector notation using its cosine components as defined in Eq.~\ref{eq:omega_cos}. Also recall the definition of the gradient, given in Eq.~\ref{eq:grad}.

\begin{equation} \label{eq:grad}
\nabla f(x, y, z) = \frac{\partial f(x, y, z)}{\partial x} \hat{i} + \frac{\partial f(x, y, z)}{\partial y} \hat{j} + \frac{\partial f(x, y, z)}{\partial z} \hat{k}
\end{equation}

To a first order approximation, a partial derivitive can be computed using Eq.~\ref{eq:deriv_1}.

\begin{equation} \label{eq:deriv_1}
\frac{\partial f(x)}{\partial x} \approx \frac{f(x+\Delta x/2) - f(x + \Delta x/2)}{\Delta x}
\end{equation}

Using the definition of $\hat{\Omega}$ and Eq.~\ref{eq:grad} the $\hat{\Omega} \cdot \nabla \psi$ term can be rewritten as

\begin{equation} \label{eq:spatial_1}
\begin{split}
\Omega \cdot \nabla \psi 
& = <\mu, \eta, \xi> \cdot <\frac{\partial \psi}{\partial x}, \frac{\partial \psi}{\partial y}, \frac{\partial \psi}{\partial z}> \\
& =
\frac{\partial \psi}{\partial x}\mu + \frac{\partial \psi}{\partial y}\eta + \frac{\partial \psi}{\partial z}\xi \\
& \approx 
\frac{\psi(x + \Delta x/2, y, z) - \psi(x - \Delta x/2, y, z)}{\Delta x} \mu \\
&+ 
\frac{\psi(x, y + \Delta y/2, z) - \psi(x, y - \Delta y/2, z)}{\Delta y} \eta \\
&+ 
\frac{\psi(x, y, z + \Delta z/2) - \psi(x, y, z - \Delta z/2)}{\Delta z} \xi
\end{split}
\end{equation}

As a particle enters the voxel shown in Fig.~\ref{fig:gradient} along some direction, $\Omega$, as indicated in Fig.~\ref{fig:gradient}, SOMETHING.

The surface flux. WRT the direction omega. This allows generalization.

\begin{figure}[tb]
  \begin{center}
   \includegraphics[width=3.75in]{figs/gradient}
  \end{center}
  \caption{The caption.}
\label{fig:gradient}
\end{figure}%

The generalization of Eq.~\ref{eq:grad_y2} is applied to Eq.~\ref{eq:deriv_1} to yield Eq.~\ref{eq:deriv_2} which is generally valid for all octants.

\begin{equation} \label{eq:deriv_2}
\Omega \cdot \nabla \psi \approx 
\frac{\psi_{x,in} - \psi_{x,out}}{\Delta x} \mu + 
\frac{\psi_{y,in} - \psi_{y,out}}{\Delta y} \eta + 
\frac{\psi_{z,in} - \psi_{z,out}}{\Delta z} \xi
\end{equation}

Using the approximation given in Eq.~\ref{eq:deriv_2}, Eq.~\ref{eq:boltz_i} can be rewritten as Eq.~\ref{eq:boltz_i2}.

\begin{equation} \label{eq:boltz_i2}
\begin{split}
&\left[ 
\frac{\psi_{i,a,x,in}^g - \psi_{i,a,x,out}^g}{\Delta x} \mu + 
\frac{\psi_{i,a,y,in}^g - \psi_{i,a,y,out}^g}{\Delta y} \eta + 
\frac{\psi_{i,a,z,in}^g - \psi_{i,a,z,out}^g}{\Delta z} \xi
\right]
+ \Sigma_{t,i}^g \psi_{i,a}^{g} \\
& = 
\sum_{a=0}^{N_a-1} \sum_{g'=g}^{G-1} \Sigma_{s, i, a, a'}^{g, g'} \psi_{i, a'}^{g'} \omega_a + S_{i,a}^g
\end{split}
\end{equation}

It is useful to point out that each of $\psi_{i,a,x,in}^g$, $\psi_{i,a,x,out}^g$, $\psi_{i,a,y,in}^g$, $\psi_{i,a,y,out}^g$, $\psi_{i,a,z,in}^g$, and $\psi_{i,a,z,out}^g$ are \textit{surface} averaged flux values while $\psi_{i,a}^{g}$ is a \textit{volume} averaged flux value.

In order for the discretized LBE to be solved using Eq.~\ref{eq:boltz_i2}, the incoming flux must always be known. Either the boundary conditions or previously calculated values determine the incoming flux. Therefore, in order to compute the average flux in a cell, some relationship between the average flux and the outgoing surface fluxes must be known. These values can be related by any one of many different mechanisms, the most common of which is the diamond difference approximation.

@@

The diamond difference approximation, relates the incoming and outgoing surface fluxs to the cell averaged flux. The average flux is assumed to be the average of the two surface fluxes. This is assumed to be the case in all three spatial dimensions. This leads to Eq.~\ref{eq:dd}.

\begin{equation} \label{eq:dd}
\begin{split}
\frac{\psi_{i,a,x,in}^g + \psi_{i,a,x,out}^g}{2} &= \psi_{i,a}^{g} \\
\frac{\psi_{i,a,y,in}^g + \psi_{i,a,y,out}^g}{2} &= \psi_{i,a}^{g} \\
\frac{\psi_{i,a,z,in}^g + \psi_{i,a,z,out}^g}{2} &= \psi_{i,a}^{g}
\end{split}
\end{equation}

Multiplying both sides of Eq.~\ref{eq:dd} and subtracting twice the outgoing surface flux from each yields Eq.~\ref{eq:dd2}.

\begin{equation} \label{eq:dd2}
\begin{split}
\psi_{i,a,x,in}^g - \psi_{i,a,x,out}^g &= 2\psi_{i,a}^{g} - \psi_{i,a,x,out}^g \\
\psi_{i,a,y,in}^g - \psi_{i,a,y,out}^g &= 2\psi_{i,a}^{g} - \psi_{i,a,y,out}^g \\
\psi_{i,a,z,in}^g - \psi_{i,a,z,out}^g &= 2\psi_{i,a}^{g} - \psi_{i,a,z,out}^g
\end{split}
\end{equation}

The form of the left hand side of Eq.~\ref{eq:dd2} is the same as in Eq.~\ref{eq:boltz_i2}. This allows Eq.~\ref{eq:dd2} to be used to make a substitution in Eq.~\ref{eq:boltz_i2} that removes the dependence on the outgoing flux from discretized equation. After the substitution:

>>>>

This chapter summarizes the solution methodology employed by DOCTORS. The techniques used to compute the collided and uncollided fluxes are described in separate sections. After the flux solution methodology is explained, the methods used for source generation and flux-to-dose conversion are covered.

@@

The linear Boltzmann equation (LBE) is generally true of any particle given that outside forces (electromagnetic or gravitational) are either not present or negligible and particle-particle collisions are insignificant. Therefore, the LBE is generally not valid for charged particles or systems whose particle density is on the order of the medium they are traveling through. In such systems, particle-particle interactions may not be negligible.

@@

The steady state form of the LBE can be readily derived from simple intuition. In the steady state, particle production and removal must be equal. In any volume, three production mechanisms are present. Particles can (1) stream into the volume across a bounding surface, (2) inscatter from another energy-direction component of phase space, or (3) be produced directly by the external source. Three removal terms are also present, particles can (1) stream out of the volume of interest by crossing a bounding surface, (2) be absorbed and vanish entirely, or (3) outscatter to a energy-direction component of phase space other than the one of interest.

The downscatter and absorption removal terms are typically grouped together using
\begin{equation}
\Sigma_t = \Sigma_a + \Sigma_s
\end{equation}
where $\Sigma_t$, $\Sigma_a$, and $\Sigma_s$ are the macroscopic total, absorption, and scatter cross sections respectively. The removal of particles from interactions (scatter and absorption) per volume is then
\begin{equation}
\Sigma_t(\boldsymbol{r}, E) \psi(\boldsymbol{r}, E, \hat{\Omega})
\end{equation}
where $\psi$ is the angular flux. The production and removal streaming operators are combined as 
\begin{equation}
\hat{\Omega} \cdot \nabla \psi(\boldsymbol{r}, E, \hat{\Omega})
\end{equation}
and included as a removal term since the normal is defined pointing outward from the volume. The normal sign convention results in particles streaming out being positive and those streaming in becoming negative.

The inscatter to a particular energy and direction is computed by integrating over all other energies and directions
\begin{equation}
\int_{4\pi}^{} \int_{0}^{\infty} \Sigma_s(\boldsymbol{r}, E' \rightarrow E, \hat{\Omega}' \rightarrow \hat{\Omega}) \psi(\boldsymbol{r}, E', \hat{\Omega}') dE' d\hat{\Omega}'.
\end{equation}
An external source $S$ produces particles per volume. Combining the removal terms on the left and the production terms on the right yields
\begin{equation} \label{eq:boltz}
\begin{split}
	&\left[ \hat{\Omega} \cdot \nabla + \Sigma_t(\boldsymbol{r}, E) \right]
	\psi(\boldsymbol{r}, E, \hat{\Omega}) = \\
	&\int_{4 \pi} \int_0^\infty \Sigma_s(\boldsymbol{r}, E' \rightarrow E, \hat{\Omega}' \rightarrow \hat{\Omega}) \psi(\boldsymbol{r}, E', \hat{\Omega}') dE' d\hat{\Omega}' + S(\boldsymbol{r}, E, \hat{\Omega})
\end{split}
\end{equation}
which is the LBE. The following sections show how Eq.~\ref{eq:boltz} is discretized and solved computationally.

\subsection{Angular Discretization}
The coordinate system used in DOCTORS is shown in Fig.~\ref{fig:coord_sys}. A discrete set of $N_a$ angles ($\Omega_{a}, \quad a = 0 \ldots N_a-1$) is selected to represent continuous directional space. Particles are transported only along these discrete directions from each voxel to adjacent voxels.

@@

THEORY AND MATH?

To include a user defined quadrature, the plain text file should be formatted such that it contains four columns similar to Table~\ref{tab:quad_format} (without any headings). No checks are performed to ensure that octants are covered, this allows simple testing quadratures that utilize only a single direction to be used. Values for $\mu$, $\eta$, and $\xi$ do not necessarily need to add to unity as they will be automatically normalized by DOCTORS. Likewise, the weights will be normalized such that their summation is unity. However, user defined directions \textit{should not} be parallel to any major axis. Doing so may result in undefined behavior. Also, after normalization, no two rows should have identical values for $\mu$, $\eta$, and $\xi$ such as the last two rows of Table~\ref{tab:quad_interp}, nor may any weights be negative. The data in Table~\ref{tab:quad_format} would be read in and renormalized by DOCTORS. The normalized data would be identical to the data in Table~\ref{tab:quad_interp}.

@@

@@

A given direction, $\hat{\Omega}$, is determined by its three cosine components: 
\begin{equation} \label{eq:omega_cos}
\hat{\Omega} = \mu \hat{i} + \eta \hat{j} + \xi \hat{k}
\end{equation}
where $\hat{i}$, $\hat{j}$, and $\hat{k}$ are the unit directions in the $x$, $y$, and $z$ directions respectively. Therefore, any discrete direction $\hat{\Omega}_a$ can be expressed as $<\mu, \eta, \xi>$. The two forms will be used interchangeably.

Discretizing Eq.~\ref{eq:boltz} in angular space gives
\begin{equation} \label{eq:boltz_a}
\begin{split}
&\left[ \hat{\Omega}_a \cdot \nabla + \Sigma_t(\boldsymbol{r}, E) \right]
\psi_{a}(\boldsymbol{r}, E) = \\
&\sum_{a=0}^{N_a-1} \int_0^\infty \Sigma_{s, a, a'}(\boldsymbol{r}, E' \rightarrow E) \psi_{a'}(\boldsymbol{r}, E') dE' \omega_a + S_a(\boldsymbol{r}, E)
\end{split}
\end{equation}
where the subscript $a$ denotes the direction number and $\omega_a$ is a weight associated with that direction. The weights are designed with the angles such that once a discrete quadrature is selected, integration over continuous space becomes integration in discrete space approximated by
\begin{equation} \label{eq:disc_int}
\int_{4 \pi} f(\hat{\Omega}) d\hat{\Omega} \approx \sum_{a=0}^{N_a-1} f(\hat{\Omega}_a) \omega_a = 1.
\end{equation}

@@

Continuous energy is discretized into $G$ groups indexed from 0 to $G-1$ as illustrated in Fig.~\ref{fig:energy_groups}. Therefore, Eq.~\ref{eq:boltz_a} becomes
\begin{equation} \label{eq:boltz_e}
\left[ \hat{\Omega}_a \cdot \nabla + \Sigma_t^g(\boldsymbol{r}) \right]
\psi_{a}^{g}(\boldsymbol{r}) = 
\sum_{a=0}^{N_a-1} \sum_{g'=0}^{G-1} \Sigma_{s, a, a'}^{g, g'}(\boldsymbol{r}) \psi_{a'}^{g'}(\boldsymbol{r}) \omega_a + S_a^g(\boldsymbol{r})
\end{equation}
where the group number is indexed by superscript $g$.

@@

Once the energy group structure is selected, cross section data must be available. Group averaged cross section values from energy $E_1$ to $E_2$ are computed using Eq.~\ref{eq:groupxs} where $\sigma(E)$ is the continuous microscopic cross section. A weighting function, $f$, is used to weight some energies. The weighting function can be any of a number commonly used functions. The SCALE data libraries us a uniform function [CITE].
\begin{equation}\label{eq:groupxs}
\sigma_G = \frac{\int_{E_1}^{E_2}f(E')\sigma(E') dE'}{\int_{E_1}^{E_2} f(E') dE'}
\end{equation}

In photon based problems, upscatter is typically negligible [CITE]. By the time photons become low enough in energy to upscatter significantly, they are no longer of dosimetric concern. Removing upscatter reduces Eq.~\ref{eq:boltz_e} to
\begin{equation} \label{eq:boltz_e2}
\left[ \hat{\Omega}_a \cdot \nabla + \Sigma_t^g(\boldsymbol{r}) \right]
\psi_{a}^{g}(\boldsymbol{r}) = 
\sum_{a=0}^{N_a-1} \sum_{g'=g}^{G-1} \Sigma_{s, a, a'}^{g, g'}(\boldsymbol{r}) \psi_{a'}^{g'}(\boldsymbol{r}) \omega_a + S_a^g(\boldsymbol{r})
\end{equation}
which is very similar; only the starting point for the scatter summation changes. Computationally, though, the lack of upscatter makes a significant difference. The highest energy group can be solved independently of all those below it. Once that group is solved, the next group can also be solved and so forth. In the presence of upscatter, the highest group depends on the solution to all groups below it so all energy groups must be solved simultaneously.

The group structure cannot be arbitrarily selected by DOCOTRS, instead it is determined by the external cross section data file loaded by the user. The energy group structure should be selected appropriately for the problem. The cross sections currently included with DOCTORS are those available from the SCALE 6.2 distribution. Those cross sections are designed for light water reactor analysis and are not well suited to medical physics applications. This is currently one of the greatest limitations of DOCOTRS. The SCALE cross section data also contains neutron data which is useless for the CT dosimetry DOCTORS performs. More details of the cross section parsing and generation are included in Section~\ref{sec:xsparse} and~\ref{sec:xsgen} respectively.

@@

The problem domain is split into an evenly spaced Cartesian grid. Uniform grid spacing is not required by discrete ordinate methods though it is helpful [CITE], but is necessarily the case with CT voxel phantoms.

The problem domain for a CT voxel phantom is a regular, rectangular parallelepiped of dimension $D_x \times D_y \times D_z$, the values of which must be known based on the CT setup by the user. The mesh is partitioned into $N_x$, $N_y$, and $N_z$ evenly spaced bins along each major direction as shown in Fig.~\ref{fig:spatial_disc}. The total number of voxels, $N_V$ is easily computed from the number of bins in each major direction:
\begin{equation} \label{eq:n_v}
N_V = N_x N_y N_z.
\end{equation}
The length of a voxel along the $x$-direction is computed by 
\begin{equation} \label{eq:mesh_x}
\Delta x = \frac{D_x}{N_x}.
\end{equation}
Analogous equations apply in the $y$ and $z$ directions as well.

Voxels can be indexed in one of two ways, either by their component indices or global index. Each voxel in the mesh has three components indices, one in each major direction, $i_x$, $i_y$, and $i_z$ that range from 0 to $N_x - 1$, $N_y - 1$, or $N_z-1$. However, this makes indexing cumbersome, so they are combined into a unique global index, $i$ using Eq.~\ref{eq:indx_flat}. In general, this is known as "flattening" a matrix into a one-dimensional vector. More details about the flattening implementation used in DOCTORS is given in Section~\ref{sec:flatten}.

@@

Using global indexing discussed, the fully discretized the steady state LBE given is
\begin{equation} \label{eq:boltz_i}
\left[ \hat{\Omega}_a \cdot \nabla + \Sigma_{t,i}^g \right]
\psi_{i,a}^{g} = 
\sum_{a=0}^{N_a-1} \sum_{g'=g}^{G-1} \Sigma_{s, i, a, a'}^{g, g'} \psi_{i, a'}^{g'} \omega_a + S_{i,a}^g
\end{equation}

@@

To solve the fully discretized form of the Linear Boltzmann Equation given in Eq.~\ref{eq:boltz_i}, the gradient operator must first be computed numerically. Recall that $\Omega_a$ can be written in vector notation using its cosine components as defined in Eq.~\ref{eq:omega_cos}. Also recall the definition of the gradient:
\begin{equation} \label{eq:grad}
\nabla f(x, y, z) = \frac{\partial f(x, y, z)}{\partial x} \hat{i} + \frac{\partial f(x, y, z)}{\partial y} \hat{j} + \frac{\partial f(x, y, z)}{\partial z} \hat{k}.
\end{equation}
To a first order approximation, a partial derivative is computed as
\begin{equation} \label{eq:deriv_1}
\frac{\partial f(\zeta)}{\partial \zeta} \approx \frac{f(\zeta+\Delta \zeta/2) - f(\zeta + \Delta \zeta/2)}{\Delta \zeta}
\end{equation}
with respect to $\zeta$. Combining Eq.~\ref{eq:grad} and~\ref{eq:deriv_1}, the $\hat{\Omega} \cdot \nabla \psi$ term can be rewritten as:
\begin{equation} \label{eq:spatial_1}
\begin{split}
\Omega \cdot \nabla \psi 
& = <\mu, \eta, \xi> \cdot <\frac{\partial \psi}{\partial x}, \frac{\partial \psi}{\partial y}, \frac{\partial \psi}{\partial z}> \\
& =
\frac{\partial \psi}{\partial x}\mu + \frac{\partial \psi}{\partial y}\eta + \frac{\partial \psi}{\partial z}\xi \\
& \approx 
\frac{\psi(x + \Delta x/2, y, z) - \psi(x - \Delta x/2, y, z)}{\Delta x} \mu \\
&+ 
\frac{\psi(x, y + \Delta y/2, z) - \psi(x, y - \Delta y/2, z)}{\Delta y} \eta \\
&+ 
\frac{\psi(x, y, z + \Delta z/2) - \psi(x, y, z - \Delta z/2)}{\Delta z} \xi.
\end{split}
\end{equation}
The flux variables $\psi_{i,a,x,in}^g$, $\psi_{i,a,x,out}^g$, $\psi_{i,a,y,in}^g$, $\psi_{i,a,y,out}^g$, $\psi_{i,a,z,in}^g$, and $\psi_{i,a,z,out}^g$ are \textit{surface} averaged flux values while $\psi_{i,a}^{g}$ is a \textit{volume} averaged flux value.

The gradient is computed with respect to th edirection $\hat{\Omega}$, thus Eq.~\ref{eq:spatial_1} is valid only for the first octant ($\mu, \eta, xi > 0$). In other octants, a new gradient must be used. Rather than enumerating all eight variants, Eq.~\ref{eq:spatial_1} can be generalized to
\begin{equation} \label{eq:deriv_2}
\Omega \cdot \nabla \psi \approx 
\frac{\psi_{x,out} - \psi_{x,in}}{\Delta x} \mu + 
\frac{\psi_{y,out} - \psi_{y,in}}{\Delta y} \eta + 
\frac{\psi_{z,out} - \psi_{z,in}}{\Delta z} \xi
\end{equation}
where
\begin{equation}
\psi_{x,out} = 
\begin{cases}
\psi(x+\Delta x/2, y, z) - \psi(x-\Delta x/2, y, z) \,, \quad \mu > 0 \\
\psi(x-\Delta x/2, y, z) - \psi(x+\Delta x/2, y, z) \,, \quad \mu < 0
\end{cases}
\end{equation}
and
\begin{equation}
\psi_{x,in} = 
\begin{cases}
\psi(x-\Delta x/2, y, z) - \psi(x+\Delta x/2, y, z) \,, \quad \mu > 0 \\
\psi(x+\Delta x/2, y, z) - \psi(x-\Delta x/2, y, z) \,, \quad \mu < 0
\end{cases}
\end{equation}
and $\psi_{y,out}$, $\psi_{y,in}$, $\psi_{z,out}$, and $\psi_{z,in}$ are similarly defined but with respect to $\eta$ and $\xi$ respectively.

%\begin{figure}[tb]
%  \begin{center}
%   \includegraphics[width=3.75in]{figs/gradient}
%  \end{center}
%  \caption{The caption.}
%\label{fig:gradient}
%\end{figure}%

Using the generalized approximation given in Eq.~\ref{eq:deriv_2}, Eq.~\ref{eq:boltz_i} can be rewritten as
\begin{equation} \label{eq:boltz_i2}
\begin{split}
&\left[ 
\frac{\psi_{i,a,x,out}^g - \psi_{i,a,x,in}^g}{\Delta x} \mu + 
\frac{\psi_{i,a,y,out}^g - \psi_{i,a,y,in}^g}{\Delta y} \eta + 
\frac{\psi_{i,a,z,out}^g - \psi_{i,a,z,in}^g}{\Delta z} \xi
\right]
+ \Sigma_{t,i}^g \psi_{i,a}^{g} \\
& = 
\sum_{a=0}^{N_a-1} \sum_{g'=g}^{G-1} \Sigma_{s, i, a, a'}^{g, g'} \psi_{i, a'}^{g'} \omega_a + S_{i,a}^g.
\end{split}
\end{equation}

In order for the discretized LBE to be solved using Eq.~\ref{eq:boltz_i2}, both the incoming and outgoing flux must be known. Therefore, in order to compute the average flux in a cell, some relationship between the average flux and the outgoing surface fluxes must be known. These values can be related by any one of many different mechanisms, the most common of which is the diamond difference approximation.

@@

The diamond difference approximation, relates the incoming and outgoing surface fluxes to the cell averaged flux. The cell averaged flux is assumed to be the average of any two opposite surface fluxes. This is assumed to be the case in all three spatial dimensions. This leads to:
\begin{equation} \label{eq:dd}
\begin{split}
\frac{\psi_{i,a,x,in}^g + \psi_{i,a,x,out}^g}{2} &= \psi_{i,a}^{g} \\
\frac{\psi_{i,a,y,in}^g + \psi_{i,a,y,out}^g}{2} &= \psi_{i,a}^{g} \\
\frac{\psi_{i,a,z,in}^g + \psi_{i,a,z,out}^g}{2} &= \psi_{i,a}^{g}.
\end{split}
\end{equation}
Multiplying both sides of Eq.~\ref{eq:dd} by 2 and subtracting twice the incoming surface flux from each yields:
\begin{equation} \label{eq:dd2}
\begin{split}
\psi_{i,a,x,out}^g - \psi_{i,a,x,in}^g &= 2\psi_{i,a}^{g} - 2\psi_{i,a,x,in}^g \\
\psi_{i,a,y,out}^g - \psi_{i,a,y,in}^g &= 2\psi_{i,a}^{g} - 2\psi_{i,a,y,in}^g \\
\psi_{i,a,z,out}^g - \psi_{i,a,z,in}^g &= 2\psi_{i,a}^{g} - 2\psi_{i,a,z,in}^g.
\end{split}
\end{equation}
